
\documentclass{book}

\usepackage{a4wide}
\usepackage{amssymb}
\usepackage{color}
\usepackage{fancyvrb}
\usepackage{framed}
\usepackage{graphicx}
\usepackage{relsize}
\usepackage{subfig}
\usepackage{url}
\usepackage{xspace}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}

\input{definitions.tex}



\begin{document}

\title{Runtime manual}
\author{Koen V. Hindriks and Wouter Pasman}
\maketitle

%
%
%
\setcounter{tocdepth}{3}




%************************ STAND ALONE RUNNING ***************************
%
%
%
\chapter{Running in Stand-Alone Mode}\label{standalone}
%
This chapter explains how to use the \GOAL runtime to run GOAL in stand-alone mode (without GUI).
This way you can use {\GOAL} from scripting
languages like a shell script or bat script, or from your Java program.

Stand-alone running always involves starting up a Java Virtual Machine (JVM)
with the right environment settings. We first describe the required environment
and then describe how to use this from a Java program or from a script.

We assume you have downloaded the latest Runtime jar from \url{https://github.com/goalhub/runtime/releases}.  


%
%
%
\section{From Java}
%
Running a stand-alone MAS from Java is very simple: just create the
goal.tools.Run object from Java and start it using this call 

\code{new Run(new File(fileName)).run();}

where fileName is the MAS file that you want to run.


%
%
%
\section{From the Command Line}
%
To run \GOAL stand-alone from the command line, the typical call looks like this:

\begin{verbatim}
java -cp runtime-...-shaded.jar goal.tools.Run <filename or directory>
\end{verbatim}

You can also use the \verb|simpleidemodules-...-shaded.jar| instead, just make sure you use the full 'shaded' jar.

\noindent where \code{<filename or directory>} is the path to the MAS file or directory with MAS files you want to run.
The tool has the following options:

\begin{tabular}{p{4cm}p{10cm}}
\hline

 \code{-d,-{}-debug}           & Display out put from debugger while running agent\\
 \code{-e}                           &Print messages from error\\
 \code{-h,--help}                 &Displays this help\\
 \code{-i}                           &Print messages from info\\
 \code{-l <dir>}                           &Create log files. Writes to given directory, or to default log directory if not set.\\
 \code{--license}               &Shows the license\\
 \code{-p}                         &Print messages from parser\\
 \code{-r,--repeats <number> }   &Number of times to repeat running all episodes\\
 \code{--recursive}           & Recursively search for mas files\\
\code{    --rmi <host>}        & Use RMI messaging middleware. Host is the location of the RMI server. Using "localhost" will initialize a RMI server\\
\code{--timeout <seconds> } & Timeout of the multi-agent system.\\
 \code{-v,--verbose }           & Print all messages\\
\code{--version}                 & Shows the current version\\
\code{ -w }                         &Print messages from warning\\
 
\hline
\end{tabular}


%
%
%
\section{Runtime Settings}
%
Some settings, e.g. used middleware, are used as set in the \code{settings.yaml} preferences file that is created after the first run in the directory where you installed the \code{runtime*.jar} file. These preferences can be changed using the GOAL IDE, by directly editing the file using a text editor, or programmatically.

Please refer to the JavaDoc for programmatically changing the various preference settings inside the
\code{goal.preferences} directory. There are preference
settings for the CorePreferences, PMPreferences, the LoggingPreferences etc.
Also notice the -w and -v options of the tool if you are looking for warning printouts.

%
%
%
\section{Batch Running}
The \code{Run} command above allows you various variants for batch running. You can specify a
directory containing multiple prepared MAS files, and/or you can request multiple runs of the MAS files. 

To enable proper batch running, your setup has to meet two requirements:

\begin{itemize}
\item Your MAS terminates at some point. This happens when all agents have terminated, e.g. exited their main program section or were killed.
\item Your environment resets the entities when it is killed and then restarted. This may seem natural but in
some configurations (eg, BW4T), killing and restarting a local environment does not necessarily reset the
real remote environment.
\end{itemize}


\bibliographystyle{plain}
\bibliography{references}

\end{document}